Dynamical properties of the Benguela Current as seen through a cold-spell algorithm.

### 2019-02-24 Write_up/Writing/Chapter1_Intro.Rmd
updated

Chapter 1: General Introduction

@@

For several decades, regime shifts have become an important factor of marine research. Regime shifts are often defined as a shift from one stable ecosystem state to another. This process affects several species within a short period of time [@cury; @jarre; @blamey). Regime shifts are thought to be driven by abiotic processes such as climate change [@polovina; @blamey]. The southern Benguela upwelling system supports a wide variety of commercially-exploited fish species and exhibits substantial variability in both oceanographic and biological components. Over the past 3 three decades, a number of species occurring within this region have undergone shifts in their distributions and Roy et al. (2007) linked these shifts to cooling waters along the inner shelf of the Agulhas bank on the south coast, east of Cape Agulhas [@blamey]. This coast is associated with high levels of harmful toxic algal blooms, which negatively impact commercial and recreational interests along the coastal region. These harmful blooms are often attributed to flagellated species with the red tide representing dinoflagellate blooms (Pitcher et al. 1993A, @pitcher1998]. Formations of red tides are often related to the prevailing winds of the southern Benguela which govern most hydrodynamic processes on the continental shelf. Red tide events are greatest during the latter part of the upwelling season during transitions in the synoptic weather patterns which result in diminished upwelling activity and increased thermal stratification (Pitcher et al. 1993a).

@@

Remotely-sensed satellite SST products are produced and maintained by a number of organisations worldwide. In order to obtain these SST data, advanced technology is launched into space, in order to obtain infra-red light being emitted by the ocean. This irradiance is then taken as a proxy for SST.  Due to the distance at which these SST are being collected, certain coarseness within the datasets are unavoidable. Satellite SST products undergo a range of quality control processes that are well researched and tested [@reynolds], making their use more reliable. However, Smit et al. (2013) have shown that SST data are not particularly reliable for coastal regions and should be used cautiously as they can be prone to both warm and cold bias. Nevertheless, satellite data products are extremely useful when investigating the ocean at broad scale and can provide important SST data.  <Will add more on the different types of data and the varying resolutions>

@@

This study made use of four satellite-derived Sea Surface Temperatures (SST) datasets and the SACTN in situ coastal SST have been observed by satellite instruments since 1981 and are regarded as one of the longest satellite-based records of any earth climate variable (REFERENCE). The National Oceanic and Atmospheric Administration (NOAA) ¼° daily Optimum Interpolation Sea Surface Temperature (OISST) is an analysis of different platforms (satellites, ships and buoys) on a global grid. This gridded daily SST product assimilates both remotely sensed and in situ sources of data to create a level-4 gap free product (Banzon et al., 2016). A spatially complete sea suface temperature map is produced by interpolation to fill in gaps within the dataset. The method used to fill these gaps includes the adjustment of satellite and ship observations which compensate for the bias differences (REFERENCE). There are two types of daily OISST, named after the relevant satellite sensors. These are the Advanced Very High Resolution Radiometer (AVHRR) and Advanced Microwave Scanning Radiometer on the Earth Observing System (AMSR-E). Currently the AVHRR has the longest record starting from 1981 to the present. Infrared instruments, like AVHRR, can make observations at relatively high resolutions. Microwave instruments, like AMSR-E, can measure SSTs in most weather conditions but not adjacent to land. The AVHRR-only Optimally-Interpolated Sea Surface. Temperature (OISST) was used to determine SST within the study region. The AVHRR datasets have been provided global SSTs for more than four decades (Reynolds and Smith, 1994; Pearce et al 2006).

@@

The Moderate-resolution Imaging Spectroradiometer (MODIS) is a scientific instrument (radiometer) on board the NASA Terra and Aqua satellite platforms with a spatial resolution in 0.041° longitude-latitude coordinates. MODIS was launched in 1991 and 2002 to study global dynamics of the earths atmosphere, ice and ocean. MODIS captures data in 36 spectral bands ranging in wavelength from 0.4 um to 14.4 um and at varying spatial resolutions (2 bands at 250 m, 5 bands at 500 m and 29 bands at 1 km), providing complete global coverage of the Earth every one to two days. MODIS Terra Global level 3 mapped thermal SST products consists of sea surface temperature data derived from the 11 and 12um thermal IR infrared (IR) Bands.

@@

This Global 1 km SST (G1SST) analysis, with a 0.01° resolution, uses satellite data from sensors that include the Advanced Very High Resolution Radiometer (AVHRR), the Advanced Along Track Scanning Radiometer (AATSR), the Spinning Enhanced Visible and Infrared Imager (SEVIRI), the Advanced Microwave Scanning Radiometer-EOS (AMSRE), the Tropical Rainfall Measuring Mission Microwave Imager (TMI), the Moderate Resolution Imaging Spectroradiometer (MODIS), the Geostationary Operational Environmental Satellite (GOES) Imager, the Multi-Functional Transport Satellite 1R (MTSAT-1R) radiometer, and in situ data from drifting and moored buoys.  Thid dataset produces a product that combines satellite data with the model output.

@@

As a result of climate change, research studies provide significant evidence relating to the increase in the warming of the ocean surface. This warming slowly propagates down through the water column and in turn increases the thermal stratification [@gruber]. Strong stratification limits the depth at which water is upwelled and hence controls the amount of nutrients brought up [@chhak; @jacox2011; @jocox2016]. The interplay between upwelling winds and stratification determines the amount of nutrients present within the system and is therfore strongly linked to biological productivity. Globally, as a result of climate change, stratification of the water-column between 0 and 200 m has increased by an estimated 4% between 1971 to 2010 [@stocker]. Increased ocean stratification as a result of surface warming may reduce the ventilation of deep-water masses. As the deep water is being up welled to the surface, decreases in oxygen and pH levels are predicted along with an increase in nutrient levels [@rykaczewski].  Generally, regional ocean climate oscillations like the Benguela Niño for the Benguela Current System impose considerable variability on temperatures and complicate attempts to assess long-term trends [@demarcq]. It is predicted that these climate oscillations may increase the variance with climate change [@timmermann; @kuzmina; @sydeman] which will further result in unusual trends.

@@

Climate change similarly affects coastal species such as oysters, corals and, sea grass. Lord (2017) also found evidence that suggests that a change in climate as a result of rising sea temperatures may increase the spread of invasive species and increase the chances of an outbreak of diseases [@altizer]. Research by Mead et al (2013) found that coastal ecosystems are at high risk as a result of climate change, later, Whitfield et al (2016) confirmed that coastal bio-diversity have been observed to be affected. Given the importance of primary productivity, it was determined that the Benguela upwelling systerm within the South African region should become the focus of this thesis. Here the increase in global temperatures are one of the main focus points along with the chlorophyll concentration. By making use of these two measurements, I am able to analyses how chlorophyll concentration vary with a change in temperature.

@@

With global warming continuing at a rapid pace, the amount of greenhouse gases within the atmosphere are going to increase in the future. This increase will undoubtedly have an impact on the world’s oceans and the marine life within, both directly and indirectly. The oceanic process of coastal upwelling is one aspect that is likely to be affected as changes in global temperature will affect temperature and pressure gradients, and in turn affect alongshore winds which are responsible for driving this process within EBUS regions. Because upwelling is important for productivity within coastal systems, a decrease in this process will negatively impact on marine fauna and flora, and on human activities such as fishing. To assess changes in upwelling events, here I aim to use SST, wind, and chlorophyll data to monitor the durations and intensities of upwelling events within the Benguela EBUS and identify wether this resulted in an extreme anomalous cold event. Atmospheric-ocean coupling is known to be affected by climate change, likely changing wind patterns and leading to upwelling favourable conditions (García-Reyes et al., 2015). Here we test if an increase in upwelling may lead to a detection of more MCSs simply because these cold temperatures caused by aseasonal upwelling will reflect in historic climatologies.

@@

This study focusses on the Benguela EBUS. The Benguela EBUS is unique in that it is bounded at both the equator ward and pole ward ends by warm water regimes. These currents are forced locally by the wind stress field off South-West Africa [@nelson ; @fennel]. This upwelling system stretches along the south-western coast of Africa from southern Angola to Cape Agulhas [@cole, 1999). Water is coldest and most persistent off the Namibian coast at 25°S and weakest at 17°S [@Lutjeharms]. The main upwelling period in the Benguela north of 25°S extends mainly from March to November, peaking during August [@shannon1985], but upwelling persists throughout the year. Upwelling season in the northern Benguela extends between September and March, with very little upwelling taking place during winter (June-August). In the central Benguela and also at Cape Frio upwelling is continuous throughout the year [@shannon1985]. The cold Benguela waters are bound in the south by the warm Agulus retroflexion region and in the north by the southward flowing Angola current. The flow of the Benguela is often considered to be topographically guided [@nelson; @barange; @barange]. The Benguela system is split into northern and southern systems by a zone of intense perennial upwelling activity in the Luderitz region within the Namibian region [@shannon; @cole].

Upwelling in the northern Benguela is in phase with the seasonal insulation cycle, whereas in the southern Benguela it is not. The result is that there is a pronounced seasonal cycle in SST in the north while in the south there is no clear SST cycle.  The seasonal warming occurring during the late summer season off northern and central Namibia is a regular occurrence and is associated with warmer, higher salinity water of Angolan origin into the region [@o; @shannon]. There have been a number of warm and cool periods documented in the northern and southern Benguela [e.g. @stander; @walker; @shannon1984; @boyd; @brainard]. Walter (1937) found a major climatic event in the northern Benguela region of anonymously warm water. This unusual warm water may have resulted due to reduced upwelling or an increase southward advection of warm water. Another warming event occurred during the summer and autumn months of 1984 [@shannon1986]. The presence of this anomalous warm water during 1984 followed two cooler than normal years in the northern Benguela. Shannon et al. (1986) illustrated the warm (21-23°C) water near Walvis Bay as a large warm water patch offshore west of Luderitz. The absence of upwelling between Walvis Bay and north of Luderitz is evident however, the upwelling centre in the Benguela system [@shannon1986], and upwelling was minimal. Interestingly, in the southern Benguela region, there was no evidence of any unusual patterns in temperature distributions. The seasonal temperature pattern within the northern Benguela is somewhat similar to that in the eastern equatorial Atlantic and the regular southward intrusion of warm water into the northern Benguela during the summer season. During these warming periods the equator ward wind stress over the northern Benguela region was stronger than normal; yet upwelling was suppressed by the southward movement of warm saline waters from the Angolan region.

Northward, the Benguela Current Upwelling system is linked by the Angola Benguela Front which converge between the Benguela current and the Benguela coastal current as well as the warm saline waters of the Angola Current [@peterson]. Research done by Yamagata and Iizuka (1995) discovered slight indication that coastal Kelvin waves, originating from equatorial Kelvin waves, carry the down welling process southward.  Southward, the coastal upwelling system is bounded by the Agulhas Bank. South of the African coast the warm Agulhas Current, western boundary current, flows in from the Indian Ocean and returns eastwards while generating warm eddies that spin off in a north-westerly direction in the Benguela current [@duncombe]. The region surrounding the Benguela experiences a persistent along shore wind. This wind is associated with the St. Helena high pressure system. Lutjeharms and Meeuwis, (1987) also found a strong relationship between the intensity of upwelling events and the strength and direction of prevailing winds. These favourable upwelling equatorial winds represent a maximum at 25°S and decrease towards the northern and southern boundaries of the Benguela upwelling system. Continental shelf bathymetry and upwelling favourable winds provide a large-scale upwelling mechanism in the southern Benguela, whereas local topography and meteorology create an alternating pattern of active and passive upwelling circulations along the coast. Winds occurring within the southern regions represent a seasonal variation, reaching a maximum during spring and summer months [@boyd; @shannon1996; @fennel]. Winds occurring within this region tend to show increasing patterns away from the coast whereas, winds occurring in the northern region exhibits relatively less seasonal variation [@shannon1985; @pitcher]. Upwelling occurring along the South Coast of the Cape Province within South Africa, specifically along the coastline is generated by local wind.  The wind increases somewhat away from the coast. This facilitated the idea of wind stress curl, as discussed in Bakun and Nelson (1991) and Shannon and Nelson (1996).

Benguela Niños are defined as anomalous warm events occurring between the southward flowing Angola current and the Benguela upwelling system off south western Africa [@shannon1986]. These warm events profoundly impact on fisheries and the climate of this region. They are proven to induce unexpected rainfall events and drastically influence fish abundance and distribution [@rouault; @boyer]. Evidently, Benguela Niños are expressed as regions of abnormal, persistent high sea surface temperatures. Many researchers often associate the equatorial interannual variability pattern in the Atlantic to the ENSO phenomenon found within the Pacific region [@sutton; @zebiak]. Benguela Niños have been recorded to occur in 1934, 1949, 1963, 1984 and 1995 [@shannon; @gammelsrod]. Picaut (1985) suggested that eastern tropic oceans are governed by remote wind stress effects through equatorial wave dynamics. Anomalies in trade winds may result in Kelvin waves that propagate eastward along the equator, inducing a deepening or a lifting of the thermocline. It is possible that equatorial Kelvin waves may play a role in the evolution of Benguela Niños. --- Need to make changes to this paragraph

@@

The warming of the global climate as a result of anthropogenic activity, has negatively influenced both terrestrial and marine ecosystems, and ultimately inhibits humanity and ecological functioning [@mcCauley]. This climate variability, manifested as increases in frequency and duration of extreme events is regarded as a great threat [@thompsonet]. Very few studies have addressed the occurrence of cold temperature spells or cold events within the Benguela region. This investigation examines recent trends and variation in the frequency and duration of cold spells occurring in this region. Due to the lack of in situ collected data, cold spells are mainly examined between the period 2014-2017 for four sites along the South African coastline falling within the Benguela upwelling cell. Here I aim to assess whether the intensity and duration of these extreme events may show any changes in frequency and intensity on a seasonal basis. In this research project I make use of various satellite derived observations of sea surface temperature (SST) and wind stress to further investigate the Benguela upwelling system along the South African coastline, whilst considering cold events and warm events occurring within this region

```{r}
wc <- site_list %>% 
  filter(coast == "wc")
save(wc, file = "wc.csv")
```

@@

>>>>

# Chapter 1: Introduction

@@

For several decades, regime shifts have become an important factor of marine research. Regime shifts are often defined as a shift from one stable ecosystem state to another. This process affects several species within a short period of time [@cury; @jarre; @blamey]. Regime shifts are thought to be driven by abiotic processes such as climate change [@polovina; @blamey]. The southern Benguela upwelling system supports a wide variety of commercially-exploited fish species and exhibits substantial variability in both oceanographic and biological components. Over the past 3 three decades, a number of species occurring within this region have undergone shifts in their distributions and Roy et al. (2007) linked these shifts to cooling waters along the inner shelf of the Agulhas bank on the south coast, east of Cape Agulhas [@blamey]. This coast is associated with high levels of harmful toxic algal blooms, which negatively impact commercial and recreational interests along the coastal region. These harmful blooms are often attributed to flagellated species with the red tide representing dinoflagellate blooms (Pitcher et al. 1993A, @pitcher1998]. Formations of red tides are often related to the prevailing winds of the southern Benguela which govern most hydrodynamic processes on the continental shelf. Red tide events are greatest during the latter part of the upwelling season during transitions in the synoptic weather patterns which result in diminished upwelling activity and increased thermal stratification (Pitcher et al. 1993a).

@@

Remotely-sensed satellite SST products are produced and maintained by a number of organisations worldwide. In order to obtain these SST data, advanced technology is launched into space, in order to obtain infra-red light being emitted by the ocean. This irradiance is then taken as a proxy for SST.  Due to the distance at which these SST are being collected, certain coarseness within the datasets are unavoidable. Satellite SST products undergo a range of quality control processes that are well researched and tested [@reynolds], making their use more reliable. However, Smit et al. (2013) have shown that SST data are not particularly reliable for coastal regions and should be used cautiously as they can be prone to both warm and cold bias. Nevertheless, satellite data products are extremely useful when investigating the ocean at broad scale and can provide important SST data.

@@

Complete datasets are often difficult to

This study made use of various satellite-derived Sea Surface Temperatures (SST) datasets. One of these datasets are the National Oceanic and Atmospheric Administration (NOAA) ¼° daily Optimum Interpolation Sea Surface Temperature (OISST). This dataset is an analysis of seawater temperature by different platforms (satellites, ships and buoys) on a global grid. The gridded daily SST product assimilates both remotely sensed and *in situ* sources of data to create a level-4 gap free product (Banzon et al., 2016). A spatially complete sea suface temperature map is produced by interpolation to fill in gaps within the dataset. The method used to fill these gaps includes the adjustment of satellite and ship observations which compensate for the bias differences (NOAA.GOV, https://www.ncdc.noaa.gov/oisst). There are two types of daily OISST, named after the relevant satellite sensors. These are the Advanced Very High Resolution Radiometer (AVHRR) and Advanced Microwave Scanning Radiometer on the Earth Observing System (AMSR-E). Currently the AVHRR has the longest record starting from 1981 to the present. Infrared instruments, like AVHRR, can make observations at relatively high resolutions. Microwave instruments, like AMSR-E, can measure SSTs in most weather conditions but not adjacent to land. The AVHRR-only Optimally-Interpolated Sea Surface. Temperature (OISST) was used to determine SST within the study region. The AVHRR datasets have been provided global SSTs for more than four decades [@reynolds; @Pearce; @corlett].

@@

The Moderate-resolution Imaging Spectroradiometer (MODIS) is a scientific instrument (radiometer) on board the NASA Terra and Aqua satellite platforms with a spatial resolution in 0.041° longitude-latitude coordinates. MODIS was launched in 1991 and 2002 to study global dynamics of the earths atmosphere, ice and ocean. MODIS captures data in 36 spectral bands ranging in wavelength from 0.4 um to 14.4 um and at varying spatial resolutions (2 bands at 250 m, 5 bands at 500 m and 29 bands at 1 km), providing complete global coverage of the Earth every one to two days [@hosoda; @gentemann, @dong]. MODIS Terra Global level 3 mapped thermal SST products consists of sea surface temperature data derived from the 11 and 12um thermal IR infrared (IR) Bands (NASA.GOV, http://oceancolor.gsfc.nasa.gov).

@@

The fifth dataset, the Global 1 km SST (G1SST) analysis, representing a 0.01° resolution, uses satellite data from sensors that include the Advanced Very High Resolution Radiometer (AVHRR), the Advanced Along Track Scanning Radiometer (AATSR), the Spinning Enhanced Visible and Infrared Imager (SEVIRI), the Advanced Microwave Scanning Radiometer-EOS (AMSRE), the Tropical Rainfall Measuring Mission Microwave Imager (TMI), the Moderate Resolution Imaging Spectroradiometer (MODIS), the Geostationary Operational Environmental Satellite (GOES) Imager, the Multi-Functional Transport Satellite 1R (MTSAT-1R) radiometer, and in situ data from drifting and moored buoys [@dash}.  Thid dataset produces a product that combines satellite data with the model output (PO.DAAC, https://podaac.jpl.nasa.gov/).

@@

As a result of climate change, research studies provide significant evidence relating to the increase in the warming of the ocean surface. This warming slowly propagates down through the water column and in turn increases the thermal stratification [@gruber]. Strong stratification limits the depth at which water is upwelled and hence controls the amount of nutrients brought up [@chhak; @jacox2011; @jocox2016]. The interplay between upwelling winds and stratification determines the amount of nutrients present within the system and is therfore strongly linked to biological productivity. Globally, as a result of climate change, stratification of the water-column between 0 and 200 m has increased by an estimated 4% between 1971 to 2010 [@stocker]. Increased ocean stratification as a result of surface warming may reduce the ventilation of deep-water masses. As the deep water is being upwelled to the surface, decreases in oxygen and pH levels are predicted along with an increase in nutrient levels [@rykaczewski].  Generally, regional ocean climate oscillations like the Benguela Niño for the Benguela Current System impose considerable variability on temperatures and complicate attempts to assess long-term trends [@demarcq]. It is predicted that these climate oscillations may increase the variance with climate change [@timmermann; @kuzmina; @sydeman] which will further result in unusual trends.

@@

Climate change similarly affects coastal species such as oysters, corals and, sea grass. Lord (2017) also found evidence that suggests that a change in climate as a result of rising sea temperatures may increase the spread of invasive species and increase the chances of an outbreak of diseases [@altizer]. Research by Mead et al (2013) found that coastal ecosystems are at high risk as a result of climate change, later, Whitfield et al (2016) confirmed that coastal bio-diversity have been observed to be affected. Given the importance of primary productivity, it was determined that the Benguela upwelling system within the South African region should become the focus of this thesis. Here the increase in global temperatures are one of the main focus points along with the chlorophyll concentration. By making use of these two measurements, I am able to analyses how chlorophyll concentration vary with a change in temperature.

@@

With global warming continuing at a rapid pace, the amount of greenhouse gases within the atmosphere are going to increase in the future. This increase will undoubtedly have an impact on the world’s oceans and the marine life within, both directly and indirectly. The oceanic process of coastal upwelling is one aspect that is likely to be affected as changes in global temperature will affect temperature and pressure gradients, and in turn affect alongshore winds which are responsible for driving the upwelling process within EBUS regions. Because upwelling is important for productivity within coastal systems, a decrease in this process will negatively impact marine fauna and flora, and human activities such as fishing. To assess changes in upwelling events, here I aim to use SST, wind, and chlorophyll data to monitor the durations and intensities of upwelling events within the Benguela EBUS and to identify wether this resulted in an extreme anomalous cold event. Atmospheric-ocean coupling is known to be affected by climate change, likely changing wind patterns and leading to upwelling favourable conditions (García-Reyes et al., 2015). Here we test if an increase in upwelling may lead to a detection of more MCSs simply because these cold temperatures caused by aseasonal upwelling will reflect in historic climatologies.

@@

The area of study for this thesis will focus on the west coast of South Africa, which is an eastern boundry upwelling system. The Benguela EBUS is unique in that it is bounded at both the equatorward and poleward ends by warm water regimes. These currents are forced locally by the wind stress field off southwest Africa [@nelson ; @fennel]. This upwelling system stretches along the south-western coast of Africa from southern Angola to Cape Agulhas [@cole]. Water is coldest and most persistent off the Namibian coast at 25°S and weakest at 17°S [@Lutjeharms]. The main upwelling period in the Benguela north of 25°S extends mainly from March to November, peaking during August [@shannon1985], but upwelling persists throughout the year. Upwelling season in the northern Benguela extends between September and March, with very little upwelling taking place during winter (June-August). In the central Benguela and also at Cape Frio upwelling is continuous throughout the year [@shannon1985]. The cold Benguela waters are bound in the south by the warm Agulus retroflexion region and in the north by the southward flowing Angola current. The flow of the Benguela is often considered to be topographically guided [@nelson; @barange; @barange]. The Benguela system is split into northern and southern systems by a zone of intense perennial upwelling activity in the Luderitz region within the Namibian region [@shannon; @cole].

Upwelling in the northern Benguela is in phase with the seasonal insulation cycle, whereas in the southern Benguela it is not. The result is that there is a pronounced seasonal cycle in SST in the north while in the south there is no clear SST cycle.  The seasonal warming occurring during the late summer season off northern and central Namibia is a regular occurrence and is associated with warmer, higher salinity water of Angolan origin into the region [@o; @shannon]. There have been a number of warm and cool periods documented in the northern and southern Benguela [e.g. @stander; @walker; @shannon1984; @boyd; @brainard]. Walter (1937) found a major climatic event in the northern Benguela region of anonymously warm water. This unusual warm water may have resulted due to reduced upwelling or an increase southward advection of warm water. Another warming event occurred during the summer and autumn months of 1984 [@shannon1986]. The presence of this anomalous warm water during 1984 followed two cooler than normal years in the northern Benguela. Shannon et al. (1986) illustrated the warm (21-23°C) water near Walvis Bay as a large warm water patch offshore west of Luderitz. The absence of upwelling between Walvis Bay and north of Luderitz is evident however, the upwelling centre in the Benguela system [@shannon1986], and upwelling was minimal. Interestingly, in the southern Benguela region, there was no evidence of any unusual patterns in temperature distributions. The seasonal temperature pattern within the northern Benguela is somewhat similar to that in the eastern equatorial Atlantic and the regular southward intrusion of warm water into the northern Benguela during the summer season. During these warming periods the equatorward wind stress over the northern Benguela region was stronger than normal; yet upwelling was suppressed by the southward movement of warm saline waters from the Angolan region.

Northward, the Benguela Current Upwelling system is linked by the Angola Benguela Front which converge between the Benguela current and the Benguela coastal current as well as the warm saline waters of the Angola Current [@peterson]. Research done by Yamagata and Iizuka (1995) discovered slight indication that coastal Kelvin waves, originating from equatorial Kelvin waves, carry the down welling process southward.  Southward, the coastal upwelling system is bounded by the Agulhas Bank. South of the African coast the warm Agulhas Current, western boundary current, flows in from the Indian Ocean and returns eastwards while generating warm eddies that spin off in a north-westerly direction in the Benguela current [@duncombe]. The region surrounding the Benguela experiences a persistent along shore wind. This wind is associated with the St. Helena high pressure system. Lutjeharms and Meeuwis, (1987) also found a strong relationship between the intensity of upwelling events and the strength and direction of prevailing winds. These favourable upwelling equatorial winds represent a maximum at 25°S and decrease towards the northern and southern boundaries of the Benguela upwelling system. Continental shelf bathymetry and upwelling favourable winds provide a large-scale upwelling mechanism in the southern Benguela, whereas local topography and meteorology create an alternating pattern of active and passive upwelling circulations along the coast. Winds occurring within the southern regions represent a seasonal variation, reaching a maximum during spring and summer months [@boyd; @shannon1996; @fennel]. Winds occurring within this region tend to show increasing patterns away from the coast whereas, winds occurring in the northern region exhibits relatively less seasonal variation [@shannon1985; @pitcher]. Upwelling occurring along the south coast of the Cape Province within South Africa, specifically along the coastline is generated by local wind. The wind increases somewhat away from the coast. This facilitated the idea of wind stress curl, as discussed in Bakun and Nelson (1991) and Shannon and Nelson (1996).

Benguela Niños are defined as anomalous warm events occurring between the southward flowing Angola current and the Benguela upwelling system off south western Africa [@shannon1986]. These warm events profoundly impact on fisheries and the climate of this region. They are proven to induce unexpected rainfall events and drastically influence fish abundance and distribution [@rouault; @boyer]. Evidently, Benguela Niños are expressed as regions of abnormal, persistent high sea surface temperatures. Many researchers often associate the equatorial interannual variability pattern in the Atlantic to the ENSO phenomenon found within the Pacific region [@sutton; @zebiak]. Benguela Niños have been recorded to occur in 1934, 1949, 1963, 1984 and 1995 [@shannon; @gammelsrod]. Picaut (1985) suggested that eastern tropic oceans are governed by remote wind stress effects through equatorial wave dynamics. Anomalies in trade winds may result in Kelvin waves that propagate eastward along the equator, inducing a deepening or a lifting of the thermocline. It is possible that equatorial Kelvin waves may play a role in the evolution of Benguela Niños. ***Need to make changes to this paragraph, random but feels important***

@@

The warming of the global climate as a result of anthropogenic activity, has negatively influenced both terrestrial and marine ecosystems, and ultimately inhibits humanity and ecological functioning [@mcCauley]. This climate variability, manifested as increases in frequency and duration of extreme events is regarded as a great threat [@thompsonet]. Very few studies have addressed the occurrence of cold temperature spells or cold events within the Benguela region. This investigation examines recent trends and variation in the frequency and duration of cold spells occurring in this region. Here I aim to assess whether the intensity and duration of these extreme events may show any changes in frequency and intensity on a seasonal basis. In this research project I make use of various satellite derived observations of sea surface temperature (SST) and wind stress to further investigate the Benguela upwelling system along the South African coastline, whilst considering cold events and warm events occurring within this region.

## Study1

## Study 2

@@

# Chapter 2: Upwelling

## Introduction

## Methods
